\section{Conclusion}
\label{sec:conclusion}

This section will conclude and summarize the results by highlighting key takeaways and putting them into the context of the scope and goals of this thesis. Finally, aspects that were not in the scope of this thesis and are subject to further research will be mentioned.

The aim of this thesis was to analyze the public transport network of Vienna for on-time performance and detect potential delay hotspots where punctuality rates are particularly low. Public transport is one of the key factors of a functioning livable city and good on-time performance is of high importance in order to ensure high usage and passenger satisfaction. In order to conduct this analysis, a custom data collection script was written which persisted 13.7 million departure records over a period of 30 days. Next, the collected data was cleaned up and filtered for incomplete or irrelevant results. For the following analysis, the data was either further filtered in order to highlight certain aspects, or it was aggregated and summarized as a whole.

Evaluating the data showed that the network offers a very good general on-time performance. While metro lines usually had very few delays due to their separation from other traffic, bus lines had both the highest and lowest punctuality rates, depending on their specific route. The initial assumption was that delays occur more often at busy interchange stations due to lots of passengers getting on and off the vehicles. This assumption did not hold in the general case, as the lowest punctuality rates were found at regular bus stations. On certain lines, however, this effect did become visible, like on tram line 71, which was analyzed in more detail in \cref{sec:line-71-analysis}. Other expected delay hotspots were terminal stations since those provide the opportunity for previously collected delays to accumulate. An interesting observation regarding this was that the terminal stops themselves surprisingly showed the best punctuality rates and even negative mean delays. However, the last 3--5 stops before them had in fact relatively low punctuality rates, suggesting that previous delays do accumulate but are however accounted for by including a leeway in the schedule. Lastly, there were some stations and even parts of whole lines that showed substantially lower punctuality rates than others. In contrast to the aforementioned cases, these however did not show any apparent systemic reason for the bad on-time performances.

In addition to the analysis done in \cref{sec:analysis}, visualizations of the delay data were presented in \cref{sec:visualization}. The aim of those visualizations was to provide an intuitive understanding of the data. The first visualization was a city map with stations plotted at their locations. Each station was colored on a gradient ranging from green to red which represents its punctuality rate. With this, the already known results from \cref{sec:analysis} were visualized as well as new delay hotspots discovered. Next, a line graph was presented which showed the number of departures and the punctuality rate per hour of the day, separated for weekdays and weekends. This showed the expected result that on-time performance drops significantly for morning and afternoon rush hour traffic during weekdays.

The presented statistics and graphics showed that there is definitely room for improvement in Vienna's public transport network, albeit the overall results suggest that there are already various measures put into practice to avoid long delays and with that unsatisfied passengers.

\input{sections/future}
